---
title: "Introduction to the OZAB Package"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(OZAB)
library(tidyverse)
library(rstan)
```

The `OZAB` package is designed to simplify the process of fitting Ordinal Zero-Augmented Beta models by combining a set of data exploration techniques and model fitting functions unique to the setting of plant cover class data. This vignette walks through three primary sections of data exploration, model fitting, and model diagnostics to showcase how the OZAB package enables the user in each stage. As an example, we will use the included example dataset, `sagebrush`, containing data on presence, absence, and abundance data for Bromus Tectorum and Artemisia Tridentata for 877 locations in the John Day Fossil Beds National Monument, a unit of the US National Park System. 

The package is designed to work seamlessly within a tidy or tidyverse framework. Therefore, many of our examples will employ the paradigms of the tidyverse including piping and `dplyr` operations.

## Data Exploration

This package provides three main unique plotting functions for data exploration of cover classes with and without covariates. The first is cumulative empirical probability plots.

### Empirical Cumulative Proportion Plots

In order to examine the empirical probabilities of each cover class with respect to a covariate, we can use the `plot_cover_class_by_covariate` function. This function takes the name of a species within the dataset and the column name of a covariate of interest. 

```{r}
sagebrush %>%
  plot_cover_class_by_covariate('Artemisia tridentata', Fire)
```

In cases where we may have a continuous response of interest, we can bin these values into meaningful or equally spaced intervals in order to produce a similar plot.

```{r}
sagebrush %>%
  mutate(Topography = cut(Topography, breaks = c(-0.25, -0.15, 0, 0.15, 0.25))) %>%
  plot_cover_class_by_covariate('Artemisia tridentata', Topography)
```

### Alluvial Plots

A unique addition to this package is the inclusion of alluvial plots which allow for an alternative visualization of the same data, perhaps more intuitively displaying the differences between covariates.

```{r}
sagebrush %>%
  filter(Species == 'Artemisia tridentata') %>%
  alluvial_plot(`Cover Class`, Fire)
```

```{r}
sagebrush %>%
  filter(Species == 'Artemisia tridentata') %>%
  mutate(Topography = cut(Topography, breaks = c(-0.25, -0.15, 0, 0.15, 0.25))) %>%
  alluvial_plot(`Cover Class`, Topography)
```

### Mosaic Plots

TBA

## Model Fitting

We can fit an OZAB model using syntax similar to `lm` in base R while using Stan to efficiently obtain posterior samples. In order to begin this example we will first generate some trasnformed parameters based on previous modeling. One important note is the `add_presence` function which is improted from the OZAB package. This function takes the cover class column assumed to be included in our dataset and appends an additioinal column consisting of boolean values where presence or abscence was reported. This function assumes that cover class values of zero indicate observed absence, but this default can be change via optional parameters. 

```{r}
### Filter Dataset, Add Presence Indicator, and Transform Covariates

sagebrush2 <-
  sagebrush %>%
    filter(Species == 'Artemisia tridentata') %>%
    add_presence(cover_class_col = `Cover Class`) %>%
    mutate(
      `Dist. to Bound` = `Dist. to Bound` / 1000,
       Topography2 = Topography^2
      )
```

In order to fit the model, we use the `ozab` function and provide formulae for Presence / Absence as well as abundance. In addition, the cutpoint scheme for the observed data must also be provided. Several convience functions have been included which hold the cutpoint vectors for commonly encountered cutpoints schemes. New or unique cutpoint schemes can be used and will be automatically validated using OZAB package functionality. All other parameters are passed to the `rstan::sampling` function to allow for granular control of estimation process. Parameters adopt the name provided via the formula with an appended prefix indicating if it is from the presence or abundance component of the model.

```{r}
### Run Model
bsage_result <-
  sagebrush2 %>%
  ozab(
    presence_formula = Presence ~ Topography + Fire + `Dist. to Bound`,
    abundance_formula = `Cover Class` ~ Topography + Topography2 + Fire,
    cutpoint_scheme = daubenmire(),
    chains = 1
    )
```

## Model Diagnostics

The resulting object from estimaiton is a `stanfit` object which allows us to use a suite of tools familiar to the user to obtain summaries and inferences. In this vignette, we will use the basic functionality provided by `rstan` to show an example of working with `stanfit` objects.

```{r}
traceplot(bsage_result)
```
